---
title: "DNA Methylation analysis"
author: "Beatriz Manso"
date: '2022-05-12'
output: 
  distill::distill_article:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

Among the epigenetic modifications, DNA methylation is important for regulating gene expression. As a gold standard technology, bisulfite sequencing provides data on genomic methylation patterns as a result of the conversion of the unmethylated cytosines into uracils that are represented as Ts in the data.


Set working directory:

```{r}
setwd("C:/Users/manso/OneDrive - University of West London/MSc Bioinformatics - UWL/6.BGA - Bioinformatics and Genome Analysis/week 8 - DNA methylation analysis")
```

Install packages and load libraries:

```{r}
if (!require("BiocManager", quietly = TRUE))
 install.packages("BiocManager")
BiocManager::install("methylKit")
BiocManager::install("genomation")

library(methylKit)
library(genomation)
```

## 1. Data and Quality measures

### Data filtering and exploratory analysis

Get Files:

```{r}
file.list=list( system.file("extdata",
"test1.myCpG.txt", package = "methylKit"),
system.file("extdata",
"test2.myCpG.txt", package = "methylKit"),
system.file("extdata",
"control1.myCpG.txt", package = "methylKit"),
system.file("extdata",
"control2.myCpG.txt", package = "methylKit") )

```



```{r}
# read the files to a methylRawList object: myobj
myobj=methRead(file.list,
sample.id=list("test1","test2","ctrl1","ctrl2"),
assembly="hg18",
treatment=c(1,1,0,0),
context="CpG"
)


## inside the methylRawList object
length(myobj)

head(myobj[[1]])

```

###  Quality check

```{r}
getMethylationStats(myobj[[2]],plot=TRUE,both.strands=FALSE)
```

**Check for general coverage statistics:**

```{r}
getCoverageStats(myobj[[2]],plot=TRUE,both.strands=FALSE)

```

**Filter:**

```{r}
filtered.myobj=filterByCoverage(myobj,lo.count=10,lo.perc=NULL,
hi.count=NULL,hi.perc=99.9)
```

### Merge samples into a single table:

**Use :: notation to make sure unite() function from methylKit is called:**

```{r}
meth=methylKit::unite(myobj, destrand=FALSE)
```

```{r}
head(meth)
```

**Create a methylBase object, where only CpGs covered with at least 1 sample per group will be returned there were two groups defined by the treatment vector, given during the creation of myobj: treatment=c(1,1,0,0)**

```{r}
meth.min=unite(myobj,min.per.group=1L)

```

### Filtering CpGs

```{r}
pm=percMethylation(meth) # get percent methylation matrix
mds=matrixStats::rowSds(pm) # calculate standard deviation of CpGs
head(meth[mds>20,])

```

```{r}
hist(mds,col="cornflowerblue",xlab="Std. dev. per CpG")

```

**methylKit object to a GRanges object to remove C->T mutations locations. These locations should be removed from the analysis as they do not represent bisulfite treatment associated conversions.**

```{r}
library(GenomicRanges)
mut=GRanges(seqnames=c("chr21","chr21"),
ranges=IRanges(start=c(9853296, 9853326),
end=c( 9853296,9853326)))

```

**Select CpGs that do not overlap with mutations:**

```{r}
sub.meth=meth[! as(meth,"GRanges") %over% mut,]
nrow(meth)

```

```{r}
nrow(sub.meth)

```

### Clustering samples

**Cluster and create dendrogram to group data points by their similarity:**

```{r}
clusterSamples(meth, dist="correlation", method="ward", plot=TRUE)
```


**Setting the plot=FALSE will return a dendrogram**

```{r}
clusterSamples(meth, dist="correlation", method="ward", plot=FALSE)
```

### PCA -  Principal component analysis

**Perform Principal component analysis (PCA) to mathematically assess possible correlated variables and uncorrelated variables:**

```{r}
PCASamples(meth, screeplot=TRUE)

```

**Plot PC1 and PC2 axis as scatter plot:**

```{r}
pc=PCASamples(meth,obj.return = TRUE, adj.lim=c(1,1))

```

## 2. Extracting interesting regions: segmentation and differential methylation

### Differential methylation with Fisher’s exact test

```{r}
getSampleID(meth)
```
```{r}
new.meth=reorganize(meth,sample.ids=c("test1","ctrl1"),treatment=c(1,0))
dmf=calculateDiffMeth(new.meth)
```

**Pool the samples from the same group by adding up the number of Cs and Ts per group:**

```{r}
pooled.meth=pool(meth,sample.ids=c("test","control"))
dm.pooledf=calculateDiffMeth(pooled.meth)
```

```{r}
#Get differentially methylated bases/regions with specific cutoffs:
all.diff=getMethylDiff(dm.pooledf,difference=25,qvalue=0.01,type="all")

#Get hyper-methylated
hyper=getMethylDiff(dm.pooledf,difference=25,qvalue=0.01,type="hyper")

#Get hypo-methylated
hypo=getMethylDiff(dm.pooledf,difference=25,qvalue=0.01,type="hypo")

#Using [ ] notation
hyper2=dm.pooledf[dm.pooledf$qvalue < 0.01 & dm.pooledf$meth.diff > 25,]

#Logistic regression based tests
dm.lr=calculateDiffMeth(meth,overdispersion = "MN",test ="Chisq")

#with Betabinomial distribution based tests
dm.dss=calculateDiffMethDSS(meth)

# Differential methylation for regions rather than base-pairs
tiles=tileMethylCounts(myobj,win.size=1000,step.size=1000)

head(tiles[[1]],3)

```

**Summarizes the methylation information over a given set of promoter regions and outputs a methylRaw or methylRawList object:**

Read the gene BED file:

```{r}
# 
gene.obj=readTranscriptFeatures(system.file("extdata", "refseq.hg18.bed.txt",
package = "methylKit"))

```

Adding covariates:

```{r}
# Adding covariates
covariates=data.frame(age=c(30,80,34,30,80,40))

sim.methylBase=dataSim(replicates=6,sites=1000,treatment=c(rep(1,3),rep(0,3)),
                       covariates=covariates,sample.ids=c(paste0("test",1:3),paste0("ctrl",1:3)))

my.diffMeth3=calculateDiffMeth(sim.methylBase, covariates=covariates,
                               overdispersion="MN",test="Chisq",mc.cores=1)
```

###  Methylation segmentation

Read Data:

```{r}
methFile=system.file("extdata","H1.chr21.chr22.rds", package="compGenomRData")
mbw=readRDS(methFile)
```

Segment the methylation data:

```{r}
res=methSeg(mbw, minSeg=10, G=1:4, join.neighbours = TRUE)
```

**Plot mean methylation value and the length of the segment as a scatter plot to examine the characteristics of the segment classes:**

```{r}
plot(res$seg.mean, log10(width(res)), pch=20,
     col=scales::alpha(rainbow(4)[as.numeric(res$seg.group)], 0.2),
     ylab="log10(length)",
     xlab="methylation proportion")


```

### Working with large files

```{r}
myobj=methRead( file.list, sample.id=list("test1","test2","ctrl1","ctrl2"),
                assembly="hg18",treatment=c(1,1,0,0), dbtype="tabix")
```

### Annotation of DMRs/DMCs and segments

Read the gene BED file:

```{r}
transcriptBED=system.file("extdata", "refseq.hg18.bed.txt",
                          package = "methylKit")
gene.obj=readTranscriptFeatures(transcriptBED)

```

**Annotate differentially methylated CpGs with promoter/exon/intron using annotation data:**

```{r}
annotateWithGeneParts(as(all.diff,"GRanges"),gene.obj)
```

**We can read the CpG island annotation and annotate our differentially methylated bases/regions with them:**

```{r}
cpg.file=system.file("extdata", "cpgi.hg18.bed.txt",
                     package = "methylKit")

cpg.obj=readFeatureFlank(cpg.file,
                         feature.flank.name=c("CpGi","shores"))
```

**Convert methylDiff object to GRanges and annotate:**

```{r}
diffCpGann=annotateWithFeatureFlank(as(all.diff,"GRanges"),
                                    cpg.obj$CpGi,cpg.obj$shores,
                                    feature.name="CpGi",flank.name="shores")
```

